 \XeTeXinputencoding "GB2312"
 \begin{abstract}
在计算流体力学研究中，各种间断问题的高分辨率数值求解方法一直是一个重要的课题。如超声速、高超声速流动中的激波，高速化学反应流动中的爆轰波，多相物质间的界面等的存在，都给数值模拟研究带来各种不同的挑战。本文正是针对求解这些问题时遇到的一些困难，发展了相关的数值求解方法。
\begin{enumerate}[(1)]
  \item
激波是可压缩流动中的一个重要现象。激波捕捉经历了半个多世纪的发展，在近年来，加权基本无振荡（WENO）格式由于既能无振荡的捕捉激波，在光滑区又具有一致高阶的精度而得到越来越多的应用。但是，已有的一些WENO格式主要是针对极值点格式精度降低的问题来进一步发展性能更好的WENO格式，如WENO-M，WENO-Z等格式。针对WENO格式在过渡点附近区域所表现出的不足，本文提出多步加权思想及多步加权基本无振荡（Multistep WENO）格式，有效提高了WENO在激波附近的计算精度。除此之外，对某些问题，如湍流的计算，光滑区域需要更高的精度来分辨高频短波，构造杂交格式是一个兼顾激波捕捉和高精度低耗散的有效手段。对于杂交格式，如何有效识别间断是关键问题，文中对比了现存的不同激波识别方法，得出基模板光滑因子（整体/局部关系）的SZ识别方法最为有效。我们还将有限紧致格式应用在更复杂的方程系统---理想磁流体（MHD）方程求解中。


\item 界面是性质不同于激波的另一类的间断问题，界面广泛的存在于如燃烧室混合、射流等流动中。除了传统的无厚度的界面以外，我们往往还需要处理超过三相以上的多相界面和存在一定界面厚度的耗散界面，耗散界面产生的原因包括不可忽略的相间扩散、相变和化学反应等。考虑到常用的界面方法，如Level-Set方法，VOF方法等在守恒性方面和非传统型界面推广中的困难，我们提出一种新型的粒子类界面方法---双信息保存（DIP）方法。该方法的主要思想是：在Euler单元内引入两类Lagrange信息点，一类限制在单元内移动，称为单元信息点，用来传信息点和单元间信息，另外一类全域追踪，称为粒子信息点，用来修正单元点更新过程中引入的误差。新的方法直接求解对流方程本身，可以很好的模拟各类界面问题。

  \item
在高速化学反应流动中，爆轰波是最常见的物理现象之一，爆轰波由诱导激波和化学反应区域构成，而化学反应是一类耗散界面问题，因此爆轰波是结合了激波和界面两类间断类型的间断问题。在爆轰波模拟中，由于反应和流动的尺度差距，如果采用简单的化学反应模型，经常会存在空间刚性问题。采用激波捕捉格式对爆轰波进行模拟时，由于反应区的数值耗散会导致源项错误的激发，可能会得到错误的爆轰波波速和非物理的结果。本文，我们将DIP方法应用在组分对流方程求解中，将反应区当成耗散界面进行捕捉，能有效避免空间刚性问题。另外，本文还将DIP方法推广到采用基元反应模型的更精细的爆轰反应求解中。数值实验表明，新的求解方法可以有效保持总组分的守恒性，避免爆轰波波前和波后组分的振荡，在总的精度方面相比于传统的求解方法同样有所提高。

\item 
  本文还针对高速化学反应流动的时间刚性问题，基于数值摄动的思想，发展了常微分方程（ODE）求解的高精度、刚性稳定、具有显形式的数值摄动（NP）格式，并将其应用在刚性化学反应流及采用了基元反应模型的爆轰波数值模拟中，达到避免迭代，提高计算效率的目的。

  \end{enumerate}


\keywords{激波捕捉格式，多步加权WENO格式，双信息保存方法，数值摄动方法，刚性化学反应流动}

\end{abstract}


\begin{englishabstract}
The numerical studies of the discontinuous problems are always one of the most important topics in the CFD field. Various  challenges come from  those discontinuities, such as, the shock waves in the supersonic and hypersonic flows, the interfaces in the multi-component flows and the detonation wave in the high speed reaction  flows. The goal of this paper is to develop corresponding numerical methods to cue problems in simulating different types of discontinuities.  

Shock waves is one of the main characters in the compressible flows, although the shock-capturing methods have been developed in nearly half a century, it is still an attractive topic.  Recent years, the weighted essantially non-oscillatory (WENO) schemes have been widely used, as they can capture the shock waves without oscillatory and get high-order accuracy in the smooth regions. However, the existing modified schemes such as WENO-Z and WENO-M, are mainly focus on improving the optimal order at critical points, but those method cannot obtain the optimal accuracy at the transition points, which connect a smooth region and a discontinuity. A new multistep method is developed and a new fifth-order WENO scheme is constructed to improve the accuracy at transition points. Besides, for some problems, such as compressible turbulance flows, higher order accuracy  is needed in the smooth region to resolve the high frequency waves. Hybrid scheme is a good idea to obtain both the advantages of shock-capturing and low dissipation in the smooth regions. The key of the hybrid method is how exactly can detect the shock waves. In this paper, we compare different existing shock detectors, the numerical results show that the SZ detector based on the property of smoothness estimators (relation between the global stencil and local stencil) is the most exact one. Further more, one kind of hybrid scheme using SZ detector---finite compact scheme is extended to solve the ideal magnetohydrodynamics (MHD) equations in this paper. 

	Interface is another kind of discontinuity which appears in the compressible multi-components flows, such as in the mixing of the  combustion chamber and jet-flows. Besides the traditional two-phases interface, in some problems we also need to solve the interface with more than three phases or with the diffusive interface model. The diffusive interface is encountered when the thickness of the interface region can not be ignored such as in the cases with diffusion, phase changing or chemical reaction.  Considering that, the existed interface methods such as Level-set method and  Volume of fluid (VOF) method have problems when they are used in the diffusive interface, in this paper, a new Euler-Lagrange method --- dual information preserving (DIP) method is constructed,  In the DIP  method, two kinds of Lagrangian points are introduced into the Eulerian mesh, one kind of points is limited in every cell called cell-point to transmit the information between the Eulerian cells and the Lagrangian  points. Another kind of point named as particle-point is traced in the whole computation to keep the accuracy and sharp ouline of the surface by correcting the information of the cell-points every time step. The new method can be used in the problems with traditional interface, multi-phases interface or diffusive interface. 

 In the high speed reacting flows, the detonation wave is a combination phenomenon of the shock and the reaction which can be treated as one kind of diffusive interface. With a simple reaction model, the difference in the scales between the reaction and convection may cause huge spatial stiff problems. Without a suitable method, a wrong detonation speed or unphysical wave structure may appear in the result.  As the DIP method can compute the diffusive interface, in this paper, DIP method is used to restrict the incorrect reaction activation caused by the numerical dissipation in the mass fraction terms.  We also develop the new method to the reacting flows with elementary reaction model to  get the conservation of components, stability near the discontinuities and more accurate results.

In this paper, we also construct a high-order  numerical perturbation (NP) method to the ordinary difference equation(ODE).  Anylysis and numerical tests show that the new method is stable for stiff ODE problems. This method is also applied in  solving the reaction Euler equations.

\englishkeywords{shock-capturing method,  multistep WENO scheme, dual information preseving method,  numerical perturbation method, stiff reacting flow}
\end{englishabstract}
